What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2022-12-28 paper/main.tex
figs 2, 6 to graphviz
port to elsarticle style

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{What's in a Name? Linear Temporal Logic Literally Represents Time Lines} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Runming Li\thanks{These authors contributed equally to this work}}{Carnegie Mellon University, USA}{runmingl@andrew.cmu.edu}{https://orcid.org/0000-0001-7600-9069}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Keerthana Gurushankar\footnotemark[1]}{Carnegie Mellon University, USA}{kgurusha@andrew.cmu.edu}{[orcid]}{}

\authorrunning{Li and Gurushankar} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Runming Li and Keerthana Gurushankar} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003766.10003770</concept_id>
       <concept_desc>Theory of computation~Automata over infinite objects</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003766.10003776</concept_id>
       <concept_desc>Theory of computation~Regular languages</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003790.10003793</concept_id>
       <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003790.10002990</concept_id>
       <concept_desc>Theory of computation~Logic and verification</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\ccsdesc[300]{Theory of computation~Regular languages}
\ccsdesc[500]{Theory of computation~Modal and temporal logics}
\ccsdesc[300]{Theory of computation~Logic and verification}

\keywords{Linear Temporal Logic, \Buchi automata, $\omega$-regular expression, model checking, visualization} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{}
\EventNoEds{1}
\EventLongTitle{}
\EventShortTitle{}
\EventAcronym{}
\EventYear{}
\EventDate{}
\EventLocation{}
\EventLogo{}
\SeriesVolume{}
\ArticleNo{1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{algorithm, algpseudocode, amsmath, amsfonts, scalerel}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,  
    urlcolor=blue,
    linkcolor=blue,
    citecolor=blue,
    pdfpagemode=FullScreen,
    }
\algrenewcommand\algorithmicrequire{\textbf{Input:}}
\algrenewcommand\algorithmicensure{\textbf{Output:}}
\crefname{algocf}{alg.}{algs.}
\Crefname{algocf}{Algorithm}{Algorithms}

@@

\newcommand{\Buchi}{B\"{u}chi }
\newcommand{\stronguntil}{\hspace{0.1cm} \mathcal{U}  \hspace{0.1cm}}
\newcommand{\strongrelease}{\hspace{0.1cm} \mathcal{M} \hspace{0.1cm}}
\newcommand{\weakuntil}{\hspace{0.1cm} \mathcal{W} \hspace{0.1cm}}
\newcommand{\weakrelease}{\hspace{0.1cm} \mathcal{R} \hspace{0.1cm}}
\newcommand{\liff}{\leftrightarrow}
\DeclareMathOperator*{\bigplus}{\scalerel*{+}{\sum}}

% hyperref must go last
\usepackage{hyperref}
% but cleveref goes "laster" than hyperref
\usepackage[capitalise,noabbrev]{cleveref}
\providecommand\algorithmname{algorithm}
\crefname{algocf}{alg.}{algs.}
\Crefname{algocf}{Algorithm}{Algorithms}

@@

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
    Linear Temporal Logic (\ltl) is widely used to specify requirements in safety-critical systems.
    However, like many formal verification techniques, it is known to be unintuitive and error-prone for human practitioners to specify and validate.
    In this paper, we provide a new timeline tool for visualizing \ltl-based specifications, which is effective at intuitively representing a wide range of formulas.
    Our tool generates timeline visualizations by translating \ltl\ formulae to intermediate representations as \Buchi automata and then $\omega$-regular expressions, and finally simplifying and visualizing the expressions.
    We provide an algorithm for this visualization, a theoretical soundness analysis, and an implementation.
\end{abstract}

@@

However, \ltl\ is a literally a language of timelines, so timeline visualizations have the potential to be a powerful tool for representing \ltl\ formulas in an intuitive manner. In this paper, we propose such a timeline visualization tool, using regular expressions, to effectively represent a wide range of \ltl-based formulas.

@@

\begin{definition}[Linear Temporal Logic (LTL)]
    % cite Kristin's survey paper
    The syntax of an LTL formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, is given by the following grammar:
    \begin{align*}
        \Phi ::= p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi
    \end{align*}\label{ltl-defn}
\end{definition}
Intuitively, $\always \Phi$ says that formula $\Phi$ is true at every time step; $\eventually \Phi$ says that formula $\Phi$ is true either now or at sometime in the future time step; and $\nextt \Phi$ says formula $\Phi$ is true at the next time step immediately after the current one.
\begin{remark}
    The definition of \ltl\ can be extended to allow release and until as shown below, but they are definable using connectives in \ref{ltl-defn}, hence they are omitted in the definition. The tool we present in this paper, however, does support those operators.
    \begin{align*}
        \Phi ::= \cdots \mid \Phi \strongrelease \Phi \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi \mid \Phi \weakuntil \Phi 
    \end{align*}
\end{remark}

@@

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$
\end{definition}
\begin{definition}[\Buchi automaton]
    A \Buchi automaton is a $5$-tuple, $G = (Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols called the alphabet $\Sigma$
        \item a partial function $\delta : Q\times \Sigma \to Q$, called the transition function
        \item an initial or start stae called $s\in Q$
        \item a set of accept states $F \subseteq Q$
    \end{enumerate}
    $G$ accepts an infinite run iff at least one of its infinitely visited states is in $F$
\end{definition}
\subsection{$\omega$-regular expression}
\begin{definition}[Regular expression]
    For word $a \in \Sigma$, regular expression is defined by the following grammar:
    \begin{align*}
        A ::= \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expression]
    Let $L(A)$ denote the language accepted by regular expression $A$. Then $L(A)$ is inductively defined as
    \begin{align*}
        L(\emptyset) & = \emptyset \\
        L(\epsilon) & = \{\epsilon\} \\
        L(a) & = \{a\} \\
        L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        L^{(0)}(A) & = \{\epsilon\} \\
        L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

@@

\begin{example}
    In figure \ref{fig:ex2}, one can reason about two timelines: the upper timeline starts with $p$ at the first time step, followed by entering the grey box with one step of $!p$ and one step of $p$. Then at the end of the grey box, we reenter the box, with the next time step being $!p$, and so on. The lower timeline starts with one step of $!p$ and one step of $p$ outside the grey box, and then enter the infinite run of $!p$ and $p$ repeating. 
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.4]{examples/ex2/ex2.gv.png}
        \caption{Example timeline for specification ``$p$ oscillates every time step''}
        \label{fig:ex2}
    \end{figure}
\end{example}

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formula into timeline visualization works as follows:
\begin{enumerate}
    \item Convert \ltl\ formula to its corresponding \Buchi automata \ref{ltl2aut}
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata \ref{aut2regex}
    \item Simplify the derived $\omega$-regular expression \ref{regex-simplify}
    \item Visualize the $\omega$-regular expression according to its structure \ref{regex2timeline}
\end{enumerate}
\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.5]{img/algo-flowchart.png}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}
\subsection{LTL to BA} \label{ltl2aut}
The translation from LTL formulas to \Buchi automata is provided by SPOT\cite{Dur22}\footnote{For our tool, we use SPOT version 2.11.}.

\subsection{BA to $\omega$-regex} \label{aut2regex}

We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_{sf}$), and for those for paths for looping from the final state back to itself (say $r_{ff}$), and finally combining those to form the $\omega$-expression for valid runs ($\bigplus_{f\in F} r_{sf}r_{ff}^\omega$). The regular expressions for finite paths are in turn found by iteratively deleting interior nodes in digraph of the automaton.

@@

\begin{algorithm}[h!]
    \caption{reduce\_dfa}
    \begin{algorithmic}
        \Require ($G, v$), a DFA with state $v$ that is not initial or final
        \Ensure Deletes state $v$ from $G$ while ensuring the same language is recognized
        \For{ every $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ }
            \If {$v$ has a self-edge  $v\xrightarrow{r_{loop}}v$}
                \State replace edges $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ with $u \xrightarrow{r_{in} r_{loop}* r_{out}} w$
            \Else
                \State replace edges $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ with $u \xrightarrow{r_{in}  r_{out}} w$
            \EndIf
        \EndFor
        %\State \Return the reduced graph $G$
    \end{algorithmic}
\end{algorithm}
\begin{algorithm}[h!]
    \caption{dfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_dfa($G, v$)
            \State combine multiedges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_{ss}| r_{sf} r_{ff}^* r_{fs})^* r_{sf} r_{ff}$
        % if called by first visit, this is r_{ss}* r_{sf}
        % if called by ba2wregex, s=f, this is r_{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}
\begin{algorithm}[h!]
    \caption{dfa2regex\_firstvisit}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression of all paths from $s$ reaching $f$ for the first time
        \State delete all out edges from $f$ in $G$
        \State \Return dfa2regex($G, s, f$)
    \end{algorithmic}
\end{algorithm}
\begin{algorithm}[h!]
    \label{alg:ba2wregex}
    \caption{ba2wregex}
    \begin{algorithmic}
        \Require $G$, a \Buchi automaton
        \Ensure The $\omega$-regular expression recognized by $G$
        \State \Return $\bigcup_{f \in F}\left({\verb|dfa2regex_firstvisit|}(G, s, f){\verb|^dfa2regex|}(G, f, f) \right) $
    \end{algorithmic}
\end{algorithm}

\subsection{$\omega$-regex simplification} \label{regex-simplify}
The $\omega$-regular expression generated in \ref{aut2regex} may not be the ``simplest'' for the purpose of visualizing the timeline. We have observed multiple patterns in the resulting $\omega$-regular expression that could be simplified. For example, an $\omega$-regular expression in the form of $r^*r^{\omega}$ represents the same timeline as $r^{\omega}$, but the latter is more intuitive. For this purpose, we devised some simplification rules in our tool, based on our observation of common patterns in the generated $\omega$-regular expression.
\paragraph*{Rule-based simplification}
Here we show a demonstrating subset of simplification rules we encoded.
\begin{alignat*}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
\end{alignat*}
\paragraph*{Result of simplification}
Those simplification rules lead to more intuitive representation of timelines. Here we demonstrate their effects using an example. 
\begin{example}
    The formula $\Phi = \always (a \limplies \eventually (\neg a))$ generates un-simplified $\omega$-regular expression $((!a) | ((aa^*(!a)))((!a) | ((aa^*(!a)))^{\omega}$, and simplified version $((!a) | ((aa^*(!a)))^{\omega}$, which corresponds to the two timelines in figure \ref{fig:unsimplified} and \ref{fig:simplified}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9-unsimplified.gv.png}
        \caption{Timeline for un-simplified $\omega$-regular expression $((!a) | ((aa^*(!a)))((!a) | ((aa^*(!a)))^{\omega}$}
        \label{fig:unsimplified}
    \end{figure}
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9.gv.png}
        \caption{Timeline for simplified $\omega$-regular expression $((!a) | ((aa^*(!a)))^{\omega}$}
        \label{fig:simplified}
    \end{figure}
\end{example}

@@

\subsection{$\omega$-regex to timeline} \label{regex2timeline}
Our tool uses Graphviz\cite{Ellson2001GraphvizO} to achieve the timeline visualization step. By construction of our algorithm, every $\omega$-regular expression is in the form of 
\[
    A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}
\]
where $A_i$ are regular expressions, and $A_{2i-1}$ could be $\epsilon$. On a high level, for each of regular expression $A_i$, we visualize it as a set of accepted input. We view each union operator as parallel timelines, and each $A_{2i-1}$ is concatenated with $A_{2i}$ repeated infinitely many times, as represented in the grey box. Figure \ref{fig:timeline} presents a generic timeline resulting from this form of construction.
\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.5]{img/timeline.png}
    \caption{Generic timeline construction from $A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}$}
    \label{fig:timeline}
\end{figure}

>>>>

\documentclass[preprint,12pt]{elsarticle}
\journal{Journal of Computer Languages}

\usepackage{amsmath, amssymb, amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}{Example}[section]
\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]

\usepackage{algorithm, algpseudocode}
\algrenewcommand\algorithmicrequire{\textbf{Input:}}
\algrenewcommand\algorithmicensure{\textbf{Output:}}

\usepackage{scalerel}
\DeclareMathOperator*{\bigplus}{\scalerel*{+}{\sum}}

\usepackage[strings]{underscore} % fixes doi underscores

\usepackage{hyperref} % brings up so many errors
\usepackage[capitalise, noabbrev]{cleveref}

@@

\newcommand{\Buchi}{B\"{u}chi }
\newcommand{\stronguntil}{\hspace{0.1cm} \mathcal{U}  \hspace{0.1cm}}
\newcommand{\strongrelease}{\hspace{0.1cm} \mathcal{M} \hspace{0.1cm}}
\newcommand{\weakuntil}{\hspace{0.1cm} \mathcal{W} \hspace{0.1cm}}
\newcommand{\weakrelease}{\hspace{0.1cm} \mathcal{R} \hspace{0.1cm}}
\newcommand{\liff}{\leftrightarrow}

@@

\begin{frontmatter}
    \title{What's in a Name? Linear Temporal Logic Literally Represents Time Lines}

    \author[inst1]{Runming Li}
    \ead{runmingl@andrew.cmu.edu}

    \author[inst1]{Keerthana Gurushankar}
    \ead{kgurusha@andrew.cmu.edu}

    \author[inst2]{Kristin Y. Rozier}
    \ead{kyrozier@iastate.edu}

    \author[inst1]{Marijn Heule \corref{cor1}}
    \ead{marijn@cmu.edu}

    \affiliation[inst1]{organization={Carnegie Mellon University},%Department and Organization
                addressline={5032 Forbes Ave}, 
                city={Pittsburgh},
                postcode={15213}, 
                state={PA},
                country={USA}}

    \affiliation[inst2]{organization={Iowa State University},%Department and Organization
                %addressline={Address Two}, 
                city={Ames},
                postcode={50010}, 
                state={IA},
                country={USA}}

    \cortext[cor1]{Corresponding author}

    \begin{abstract}
        Linear Temporal Logic (\ltl) is widely used to specify requirements in safety-critical systems.
        However, like many formal verification techniques, it is known to be unintuitive and error-prone for human practitioners to specify and validate.
        In this paper, we provide a new timeline tool for visualizing \ltl-based specifications, which is effective at intuitively representing a wide range of formulas.
        Our tool generates timeline visualizations by translating \ltl\ formulae to intermediate representations as \Buchi automata and then $\omega$-regular expressions, and finally simplifying and visualizing the expressions.
        We provide an algorithm for this visualization, a theoretical soundness analysis, and an implementation.
    \end{abstract}

    \begin{highlights} %TODO
        \item Research highlight 1
        \item Research highlight 2
    \end{highlights}

    \begin{keyword}
        Modal and Temporal Logics 
        \sep Logic and Verification
        \sep Regular languages
    \end{keyword}
\end{frontmatter}

@@

However, \ltl\ is literally a language of timelines, so timeline visualizations have the potential to be a powerful tool for representing \ltl\ formulas in an intuitive manner. In this paper, we propose such a timeline visualization tool, using regular expressions, to effectively represent a wide range of \ltl-based formulas.

@@

\begin{definition}[Linear Temporal Logic (LTL)]
    % cite Kristin's survey paper
    The syntax of an LTL formula over a set of atomic propositions $\AP$, where $p\in\AP$ is a propositional variable, is given by the following grammar:
    \begin{align*}
        \Phi ::= p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi
    \end{align*}\label{ltl-defn}
\end{definition}
Intuitively, $\always \Phi$ says that formula $\Phi$ is true at every time step; $\eventually \Phi$ says that formula $\Phi$ is true either now or at sometime in the future time step; and $\nextt \Phi$ says formula $\Phi$ is true at the next time step immediately after the current one.
\begin{remark}
    The definition of \ltl\ can be extended to allow release and until as shown below, but they are definable using connectives in \cref{ltl-defn}, hence they are omitted in the definition. The tool we present in this paper, however, does support those operators.
    \begin{align*}
        \Phi ::= \cdots \mid \Phi \strongrelease \Phi \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi \mid \Phi \weakuntil \Phi 
    \end{align*}
\end{remark}

@@

\subsection{State-based \Buchi automata (BA)}
\begin{definition}[$\omega$-word]
    An $\omega$-word or infinite run of $\Sigma$, is an infinite string $s = (s_0, s_1, s_2, \dots)$ where each $s_i\in \Sigma$
\end{definition}
\begin{definition}[\Buchi automaton]
    A \Buchi automaton is a $5$-tuple, $G = (Q, \Sigma, \delta, s, F)$ consisting of
    \begin{enumerate}
        \item a finite set of states $Q$
        \item a finite alphabet of input symbols called the alphabet $\Sigma$
        \item a partial function $\delta : Q\times \Sigma \to Q$, called the transition function
        \item an initial or start state called $s\in Q$
        \item a set of accept states $F \subseteq Q$
    \end{enumerate}
    $G$ accepts an infinite run iff at least one of its infinitely visited states is in $F$
\end{definition}

\subsection{$\omega$-regular expression}
\begin{definition}[Regular expression]
    For word $a \in \Sigma$, regular expression is defined by the following grammar:
    \begin{align*}
        A ::= \emptyset \mid \epsilon \mid a \mid AA \mid A + A \mid A^*
    \end{align*}
\end{definition}
\begin{definition}[Semantics of regular expression]
    Let $L(A)$ denote the language accepted by regular expression $A$. Then $L(A)$ is inductively defined as
    \begin{align*}
        L(\emptyset) & = \emptyset \\
        L(\epsilon) & = \{\epsilon\} \\
        L(a) & = \{a\} \\
        L(A_1A_2) & = \{s_1s_2 \mid s_1 \in L(A_1) \text{ and } s_2 \in L(A_2)\} \\
        L(A_1 + A_2) & = L(A_1) \cup L(A_2) \\
        L^{(0)}(A) & = \{\epsilon\} \\
        L^{(i + 1)}(A) & = \{s_1s_2 \mid s_1 \in L(A) \text{ and } s_2 \in L^{(i)}(A)\} \\\
        L(A^*) & = \bigcup_{i \ge 0} L^{(i)}(A)
    \end{align*}
\end{definition}

@@

\begin{example}
    In figure \ref{fig:ex2}, one can reason about two timelines: the upper timeline starts with $p$ at the first time step, followed by entering the grey box with one step of $!p$ and one step of $p$. Then at the end of the grey box, we reenter the box, with the next time step being $!p$, and so on. The lower timeline starts with one step of $!p$ and one step of $p$ outside the grey box, and then enter the infinite run of $!p$ and $p$ repeating. 
    \begin{figure}[!h]
        \centering
        \includegraphics[scale=0.4]{examples/ex2/ex2.gv.png}
        \caption{Example timeline for specification ``$p$ oscillates every time step''}
        \label{fig:ex2}
    \end{figure}
\end{example}

\section{Algorithm}
On a high level, our algorithm of converting \ltl\ formulas into timeline visualization works as follows:
\begin{enumerate}
    \item Convert the given \ltl\ formula to its corresponding \Buchi automata % (as detailed in \cref{ltl2aut})
    \item Derive the $\omega$-regular expression corresponding to the \Buchi automata % (in \cref{aut2regex})
    \item Simplify the derived $\omega$-regular expression % (in \cref{regex-simplify})
    \item Visualize the $\omega$-regular expression according to its structure %(in \cref{regex2timeline})
\end{enumerate}
\begin{figure}[h!]
    \centering
    \includegraphics[width=\textwidth, trim=1cm 3cm 1cm 3cm, clip]{img/algorithm_outline.gv.pdf}
    \caption{Algorithm outline}
    \label{fig:algo}
\end{figure}

\subsection{LTL to BA} \label{ltl2aut}
The translation from LTL formulas to \Buchi automata is provided by SPOT~\cite{Dur22}\footnote{For our tool, we use SPOT version 2.11.}.

\subsection{BA to $\omega$-regex} \label{aut2regex}
We translate \Buchi automata to $\omega$-regular expressions by first finding the regular expressions for paths from the start state to some final state (say $r_{sf}$), and for those for paths for looping from the final state back to itself (say $r_{ff}$), and finally combining those to form the $\omega$-expression for valid runs ($\bigplus_{f\in F} r_{sf}r_{ff}^\omega$). The regular expressions for finite paths are in turn found by iteratively deleting interior nodes in digraph of the automaton.

@@

\begin{algorithm}[h!]
    \caption{reduce\_dfa}
    \begin{algorithmic}
        \Require ($G, v$), a DFA with state $v$ that is not initial or final
        \Ensure Deletes state $v$ from $G$ while ensuring the same language is recognized
        \For{ every $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ }
            \If {$v$ has a self-edge  $v\xrightarrow{r_{loop}}v$}
                \State replace edges $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ with $u \xrightarrow{r_{in} r_{loop}* r_{out}} w$
            \Else
                \State replace edges $u\xrightarrow{r_{in}} v, v \xrightarrow{r_{out}} w$ with $u \xrightarrow{r_{in}  r_{out}} w$
            \EndIf
        \EndFor
        %\State \Return the reduced graph $G$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{dfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_dfa($G, v$)
            \State combine multiedges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_{ss}| r_{sf} r_{ff}^* r_{fs})^* r_{sf} r_{ff}$
        % if called by first visit, this is r_{ss}* r_{sf}
        % if called by ba2wregex, s=f, this is r_{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \caption{dfa2regex\_firstvisit}
    \begin{algorithmic}
        \Require ($G, s, f$), a DFA with initial state $s$ and final state $f$
        \Ensure The regular expression of all paths from $s$ reaching $f$ for the first time
        \State delete all out edges from $f$ in $G$
        \State \Return dfa2regex($G, s, f$)
    \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
    \label{alg:ba2wregex}
    \caption{ba2wregex}
    \begin{algorithmic}
        \Require $G$, a \Buchi automaton
        \Ensure The $\omega$-regular expression recognized by $G$
        \State \Return $\bigcup_{f \in F}\left(\text{dfa2regex\_firstvisit}(G, s, f)\text{\^{}dfa2regex}(G, f, f) \right) $
    \end{algorithmic}
\end{algorithm}

\subsection{$\omega$-regex simplification} \label{regex-simplify}

The $\omega$-regular expression generated in \cref{aut2regex} may not be the ``simplest'' for the purpose of visualizing the timeline. We have observed multiple patterns in the resulting $\omega$-regular expression that could be simplified. For example, an $\omega$-regular expression in the form of $r^*r^{\omega}$ represents the same timeline as $r^{\omega}$, but the latter is more intuitive. For this purpose, we devised some simplification rules in our tool, based on our observation of common patterns in the generated $\omega$-regular expression.

\paragraph*{Rule-based simplification}
Here we show a demonstrating subset of simplification rules we encoded.
\begin{alignat*}{2}
        & r_1 + r_1r_2^* && \Longrightarrow r_1r_2^* \\
        & r + r && \Longrightarrow r \\
        & r_1 + r_2^*r_1 && \Longrightarrow r_2^*r_1 \\
        & (r^*)^{\omega} && \Longrightarrow r^{\omega} \\
        & (r_1r_2^*)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & (r_1r_2)r_2^{\omega} && \Longrightarrow r_1r_2^{\omega} \\
        & r^*r^{\omega} && \Longrightarrow r^{\omega} \\
        & rr^{\omega} && \Longrightarrow r^{\omega}
\end{alignat*}

\paragraph*{Result of simplification}
Those simplification rules lead to more intuitive representation of timelines. Here we demonstrate their effects using an example.

\begin{example}
    The formula $\Phi = \always (a \limplies \eventually (\neg a))$ generates un-simplified $\omega$-regular expression $((!a) | ((aa^*(!a)))((!a) | ((aa^*(!a)))^{\omega}$, and simplified version $((!a) | ((aa^*(!a)))^{\omega}$, which corresponds to the two timelines in figure \cref{fig:unsimplified} and \cref{fig:simplified}.
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9-unsimplified.gv.png}
        \caption{Timeline for un-simplified $\omega$-regular expression $((!a) | ((aa^*(!a)))((!a) | ((aa^*(!a)))^{\omega}$}
        \label{fig:unsimplified}
    \end{figure}
    \begin{figure}[h!]
        \centering
        \includegraphics[scale=0.3]{examples/ex9/ex9.gv.png}
        \caption{Timeline for simplified $\omega$-regular expression $((!a) | ((aa^*(!a)))^{\omega}$}
        \label{fig:simplified}
    \end{figure}
\end{example}

@@

\subsection{$\omega$-regex to timeline} \label{regex2timeline}
Our tool uses Graphviz~\cite{Ellson2001GraphvizO} to achieve the timeline visualization step. By construction of our algorithm, every $\omega$-regular expression is in the form of 
\[
    A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}
\]
where $A_i$ are regular expressions, and $A_{2i-1}$ could be $\epsilon$. On a high level, for each of regular expression $A_i$, we visualize it as a set of accepted input. We view each union operator as parallel timelines, and each $A_{2i-1}$ is concatenated with $A_{2i}$ repeated infinitely many times, as represented in the grey box. \cref{fig:timeline} presents a generic timeline resulting from this form of construction.
\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.5]{img/timeline.png}
    \caption{Generic timeline construction from $A_1A_2^{\omega} + A_3A_4^{\omega} + \cdots + A_{2n-1}A_{2n}^{\omega}$}
    \label{fig:timeline}
\end{figure}
